Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-09 Ben_Manuscripts/hdphmm/Draft.tex
New figures, additions and organization to draft

@@

  \begin{abstract}
  %MRS: added ``in soft matter systems'' to make it clear that the focus on understanding things in physical systems.
  Appropriate time series modeling of complex diffusion in soft matter systems on the microsecond time scale
  can provide a path towards inferring transport mechanisms and predicting bulk properties
  characteristic of much longer time scales. In this work we apply the infinite hidden 
  Markov model (IHMM) to solute center-of-mass trajectories generated from long 
  molecular dynamics simulations in a cross-linked inverted hexagonal phase lyotropic 
  liquid crystal (LLC) membrane in order to automatically detect and parameterize an unknown 
  number of distinct latent vector autoregressive dynamical modes. 
  %MRS: in above sentence: try to rephrase around the physical information obtained (mechanism or properties) rather than the mathematical route to it.

  %MRS: below - it's not the solute behavior, though, it's states identified by them model. Rephrase?
  We group similar solute behavior into clusters, analyze the dynamics characteristic
  to each cluster and relate them to physical interactions between solutes and the 
  membrane. This enhancement to our understanding of the diverse range of influential 
  solute-membrane interactions allows us hypothesize design changes to LLC monomers 
  aimed towards controlling the rates of solute passage, thus improving the selective 
  performance of LLC membranes.

@@

  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms 
  %MRS: emphasizing things, not just analysis.
  in porous nanoscale materials
  and project complex diffusive behavior on long time scales.

@@

  % BJC: Not sure if this should be the first paragraph, or if I could just
  % delete this paragraph and start with the next one. 
  % MRS: going to JPC, so emphasize the physical systems.
  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams.
  \begin{itemize}
    \item Desalination and boric acid removal from seawater.
    \item Organic micropollutants
    \item Hydraulic fracturing flowback water
    \item Breathable chemical warfare agent barriers
    \item High purity chemicals
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

  The ability to efficiently design novel membrane materials for selective separations would be 
  greatly enhanced by easily interpretable methods for connecting microscopic fluctuations
  with macroscopic observables.
  %MRS: should explicitly point out that just calculating diffusion coefficients don't work for subdiffusive behavior, so
  %MRS need a different way to do it.
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	%BJC: this could potentially be moved further down
  	\item There are two main challenges associated with this goal.
  	%since complex and
  	%diverse diffusive motion on time scales accessible to molecular simulations prevents
  	%us from easily relating chemically dependent solute motion to experimental time scales.
  	\item First, we need a simple way to identify and parameterize the fluctuations
  	characteristic to dominant solute-membrane interactions.
  	\item Second, we need a way to propagate this behavior far beyond the time scales
  	afforded by MD.
  \end{itemize}

  %MRS: point out that LLC's are examples where there are lots of hopping/jumping (cite previous papers) and
  %MRS: so we need different approaches.
  In this work, we are particularly interested in the design of lyotropic liquid crystal (LLC)
  monomers, a class of amphiphilic molecules whose ordered self-assembled phases can be 
  cross-linked into mechanically strong membranes capable of highly selective separations.
  \begin{itemize}
  	\item Inverted hexagonal, H\textsubscript{II}, phase LLC membranes are characterized
  	by hexagonally packed, uniform-sized and straight pores, an ideal geometry for high 
  	throughput transport.
	\item Their pores are lined with the LLC monomer functional groups which can
	potentially be designed to interact with solutes in a chemically-specific manner.
	\item It would be extremely useful to experimentalists if we could connect LLC monomer
	design with measurable quantities.
  \end{itemize}

@@

  We aim to address two primary scientific 
  questions with this work. 
  %BJC: this may be redundant from the second paragraph. Maybe better suited here?
  First, we want
  to know if the IHMM algorithm can help researchers efficiently uncover underlying
  transport mechanisms which give rise to different solute dynamical behavior.
  %MRS: flipping for clarity.
  %Second, we want to evaluate the ability of our model to reproduce solute trajectories
  %in a stochastic manner so that we can use it to predict macroscopic transport 
  %properties which would otherwise require much longer MD simulations.
  Second, we want to predict macroscopic transport by using our model to generate 
  stochastic solute trajectories at much longer timescales than possible with MD simulations, 
  extending into timescale where diffusive behavior again occurs.

@@

%  Because independent solute trajectories generated by MD simulations share 
%  similar state dynamics, we can use the IHMM to generate stochastic trajectory 
%  realizations which sample from all behavior exhibited by MD. We verify that our
%  stochastic realizations reproduce the expected hopping and trapping behavior 
%  exhibited by MD. We quantitatively validate these trajectories by comparing 
%  their mean squared displacement (MSD) to those measured by MD.

%
%%  \noindent Amphiphilic molecules are capable of self-assembling into ordered nanostructures.
%%  \begin{itemize}
%%    \item 
%%  \end{itemize}
%
%  There are limits to what we can learn from experiment about LLC membrane design.
%  \begin{itemize}
%    \item Experimental observables like permeability and selectivity allow us to speculate
%    about the molecular origins of separation processes.
%    %MRS2: not quite sure what you mean below?
%    \item This drives an empirical design approach which can potentially neglect key
%    interactions which influence selectivity.
%    \item LLC membranes have been shown to exhibit selectivities which cannot be fully 
%    explained by relatively simple macroscopic models. 
%  \end{itemize}
%  
%  Molecular Dynamics (MD) simulations can give us mechanistic insights with atomistic 
%  resolution so that we can intelligently design new membranes for solute-specific 
%  separations.
%  \begin{itemize}
%    \item In our previous work, we built a detailed atomistic model which we used
%    to understand the nanoscopic structure of an LLC Membrane.~\cite{coscia_understanding_2019}
%    \item We also used the model in order to gain a qualitative understanding of 
%    trapping mechanisms which lead to subdiffusive transport behavior.~\cite{coscia_chemically_2019}
%  \end{itemize}
%
%  Unfortunately, the timescales that we can simulate with MD are insufficient to be
%  able to make well-converged predictions of macroscopic transport properties 
%  traditionally used to characterize membranes in the lab.
%  \begin{itemize}
%    %MRS2: would be good to get this next point into the first line of a paragraph, since it's quite important. 
%    \item However, if we use descriptive stochastic models that can capture solute
%    dynamics, then we could project long timescale behavior in addition to gaining
%    a deeper understanding of solute behavior on short timescales. 
%  \end{itemize}

@@

  %MRS: consider added subsubsections to make it more readable.
  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to the dynamical behavior which give rise to those types
  of observations. The probability of transitioning to a state based on the current
  state is mathematically defined in terms of an $n\times n$ transition probability
  matrix, $T$, where $n$ is the number of states. Unfortunately, standard HMMs 
  require $n$ to be known \textit{a priori}. One can partially overcome this by 
  testing a range of numbers of hidden states and determining which is the best 
  representation of the data.~\cite{pohle_selecting_2017}

@@

  We describe the dynamics of each state visited by solutes using a first order vector 
  autoregressive (VAR(1)) model. In general, a VAR($r$) process is characterized by a
  vector of observations in a time series that are linearly dependent on $r$ previous
  values of the time series vector:
  \begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  \label{eqn:var}
  \end{equation}
  Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  process is further characterized by a shift in the mean of each dimension by the
  vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  covariance, $\Sigma$. We limited our analysis to an autoregressive order of $r=1$.
  This means that we only parameterize $A_1$. To simplify notation, we will just
  call it $A$. We used a matrix-normal inverse-Wishart prior on parameters $A$ and 
  $\Sigma$ and a Gaussian prior on $\mathbf{c}$ in order to infer their 
  values.~\cite{fox_nonparametric_2009}

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a python implementation of this process which we heavily adapted from
  %BJC: fox_bayesian is citation Emily Fox asked me to use.
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  Therefore, we looked for convergence of each entry in the $\mathbf{c}$ vector as well
  as the $A$ and $\Sigma$ matrices of Equation~\ref{eqn:var}. We detected equilibration 
  of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} on
  the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  %BJC: This paragraph could probably be cut down quite a bit. Leave justification for SI
  We carefully applied the IHMM algorithm in a way which takes advantage of the
  system's cylindrical symmetry. While there
  %There 
  are a number of viable ways in which one could choose to analyze these time series,
  %BJC: cutting down on verbosity
  % We could use all 3 dimensions straight
  %from the output trajectory. We could use just the $z$ dimensional coordinate 
  %since we are primarily interested in through-plane membrane transport. Ultimately, 
  we chose to use a multi-step procedure that we believe adequately distinguishes 
  every type of distinct dynamical behavior exhibited by the solutes. The final
  parameterization of the model is in cylindrical coordinates since each pore is a
  cylinder where we expect solutes to exhibit radially symmetric dynamics. We 
  present a step-by-step demonstration of the procedure that follows in 
  Section~\ref{S-section:ihmm_procedure} of the Supporting Information.

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$.
  %(see Figure~\ref{fig:cartesian_cylinder}). 
  By working in Cartesian coordinates, we avoid
  mathematical complexity introduced by cylindrical coordinates while estimating the 
  state sequence.

% BJC6: I don't think this figure is necessary
%  \begin{figure}
%  \centering
%  %MRS3: does this preserve the z-axis symmetry? Is it just ensuring
%  %that you can't jump that far, or is it limiting the values that
%  %the mean can take? Apologies if it's not clear to me yet.  I'm
%  %pretty sure it's just limiting the jump (since y_t = c + f(y_t-1)),
%  %but would be good to make that explicit - ``i.e. we need to make sure it doesn't jump too far'')
%  %BJC5: preserving z-axis symmetry isn't needed yet at this stage. I use the
%  % means in z in order to distinguish states (specifically estimating the c vector).
%  % I get the radial means this way. Then just ignore z later.
%  \includegraphics[width=0.2\textwidth]{cartesian_cylinder.pdf}
%  \caption{Using the nearest pore center as the origin, we represented the solute's
%  location along the pore axis in terms of the $z$ coordinates and their radial distance
%  from the pore centers in 2 dimensions, $x$ and $y$.}\label{fig:cartesian_cylinder}
%  \end{figure}

  We applied the IHMM to each of the 24 solute trajectories independently.
  Although the IHMM is capable of identifying an infinite number of states, 
  a Dirichlet process tends to exhibit a ``rich get richer" effect, favoring
  a fewer number of states.~\cite{dreyer_discovering_2011} By applying the algorithm to each trajectory 
  independently, we reduce the possibility of lumping together multiple 
  similar states which we would prefer to stay separated before clustering.
  Note that the states identified by the IHMM are heavily influenced by 
  the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. 
  In Section~\ref{S-section:prior_guesses} of the Supporting Information we 
  outline our method for choosing reliable prior parameters. 
%  The entries of $A$ and $\Sigma$ do 
%  not vary over a wide range, so the final parameters were relatively insensitive
%  to their priors.

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences. We evaluated the convergence of the state sequence based on
  the convergence of the VAR parameters. The parameters of each state
  generally converge within 500--1000 iterations. However, the boundaries of 
  each state segment fluctuate which can lead to high variance in the $A$
  and $\Sigma$ parameters in segments with a relatively low number of emissions.
  In these cases, 
  %MRS: maybe plateauing of the mean below?  Be more precise.
  the mean in each dimension tends to be a more reliable indicator of
  convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting Information).

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) 
  %MRS: not needed.
  %because $x$ and $y$ aren't meaningful alone 
  due to the system's radial symmetry. We converted the $x$ and 
  $y$ center-of-mass coordinates to $r$. We fixed the state sequence from the 
  3D parameterization and applied the inference component of the IHMM procedure
  to the cylindrical trajectories in order to estimate the VAR(1) parameters in
  terms of $r$ and $z$. Since we fixed the state sequence, and we are using conjugate
  priors, the VAR parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
  Therefore, we only ran the inference procedure for 100 iterations. 
  %MRS: hmm. Here the new covariance parameters is redone in r - does there need to be any transformation applied? 
  %MRS: though might be OK, since x^2 + y^2 = r^2 so maybe covariance are additive if x and y uncorrelated (which they should be)
  %However, note that the uncertainty in their
  %estimates are a function of the number of data points used in their estimate.

@@

@@

  \subsection{Parameter Clustering}\label{method:clustering}

@@

  %BJC: to be reworked after I have a working procedure.
  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines. We used the silhouette 
  test in order to score the quality of clustering as a function of the number 
  of clusters chosen.~\cite{kaufman_finding_2009} For our data, the silhouette test generally favors the 
  lowest number of clusters possible (see Section~\ref{S-section:nclusters} of
  the Supporting Information). However, choosing too few clusters tends to 
  not distinguish between visually obvious differences in 
  dynamic behavior. This results in finalized parameter sets that are averages of
  distinct behavior which presents further problems with the predictive modeling
  that we discuss later on. We aimed to maintain the highest silhouette score,
  and thus lowest number of clusters, possible while verifying that visually 
  distinct states stayed separated. Based on these guidelines, we decided to 
  group the $A$ and $\Sigma$ parameters into five clusters each and the radial
  means into three clusters. Of the 75 distinct states allowed by this formulation,
  the solutes in this study showed behavior from 32--36 distinct clusters. While
  this may seem like a high number of states, many are sampled infrequently.
  %MRS: what is the consequence of being sampled infrequently.

  We remapped the state sequence based on the cluster assignments and parameterized
  the clusters using the inference component of the IHMM algorithm. 
%  generated a
%  state transition probability matrix, $T$. The IHMM algorithm also produces an estimate
%  of $T$, but since we fixed the state sequence, we decided to explicitly calculate $T$ 
%  by counting the number of transitions between states compiled across all trajectories.
%  We used the parameter inference component of the IHMM algorithm in order to 
%  infer $A$ and $\Sigma$ of the clustered states.
%  We could not simply take the mean of the clustered $A$ and $\mathbf{e}_t$ 
%  parameters because it is not clear that this is a linear 
%  operation for this problem. To circumvent this problem, 
  First, we modified the ($r$, $z$) solute trajectories so that they had a mean of zero,
  leaving only the fluctuations. We did this by subtracting the mean from each 
  same-state segment of the unclustered trajectory. We used the IHMM algorithm on this
  modified trajectory to infer the clustered state parameters by fixing the clustered 
  state sequence.

@@

  %BJC:This next paragraph can probably be moved
%  Finally, we generated stochastic trajectory realizations using the clustered
%  parameter sets. We drew state sequences with transition probabilities given by
%  $T$. While in a given state, we simulated motion according to the VAR(1)
%  parameterization of that state. After each state transition, we reset the 
%  unconditional mean of each state based on the particle's position immediately
%  before the state transition occurred.

  %MRS: maybe ``using the parameterized model to explore mechanisms''
  \subsection{Tools for Exploring Mechanisms}\label{method:interactions}

@@

  % BJC: It probably makes sense to put the trajectory generation and MSD predictions
  % first since it will be part of how we justify our clustering choices.

  Solutes in this system show a wide range of behavior influenced by the 
  heterogeneity of the membrane's nanostructure as well as the interactions 
  between monomer and solute chemical functionality. Our application of the 
  IHMM can identify and distinguish these behaviors. The following approach 
  to analysis demonstrates how one can discover and explain the very complex
  behavior exhibited by solutes over long MD trajectories in terms of 
  solute-membrane interactions. In addition, we can use the IHMM in order to generate
  stochastic trajectories, with dynamics similar to MD, so that we can estimate 
  macroscopic transport properties. 	
  %MRS: subsection lettering is really big - makes it seem like a very different section?

@@

  We applied the IHMM to all trajectories of each of the four solutes studied.
  Based on our choice of clustering, we identified over 30 distinct dynamical
  modes for each solute. Many of these modes are sampled very infrequently and thus
  do not contribute significantly to a general understanding of solute transport behavior.
  In this section, we explore a subset of this data, focusing on the states
  and dominant interactions most common to solute trajectories.

  %MRS: maybe phrase it around types of mechanism, and use methanol as an example.
  \subsubsection*{The Dynamics of Methanol}

  Methanol exhibits a range of dynamic behavior. In Figure~\ref{fig:mechanism_map}, we
  plot a representative 2D methanol trajectory in cylindrical coordinates. The coordinates
  are colored according to the cluster with which methanol's dynamics are most consistent,
  which implies that this particular methanol trajectory exhibits eight distinct dynamical 
  behaviors (in order of their first appearance, they are colored dark yellow, red, light
  blue, light green, orange, blue, teal and lime green). The solute time series are coupled
  with color coded bars which describe the solute's local environment and physical 
  interactions that it undergoes at that trajectory frame. From the top, they describe 
  whether the solute is associated with sodium, whether the solute is hydrogen bonding
  with the monomer as well as to what part of the monomer it is hydrogen bonding, and 
  finally, the local number density surrounding the solute.

>>>>

\SectionNumbersOn

@@

  \begin{abstract}

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the infinite hidden Markov model (IHMM) to solute center-of-mass trajectories
  generated from long molecular dynamics simulations in a cross-linked inverted hexagonal
  phase lyotropic liquid crystal (LLC) membrane in order to automatically detect a
  variety of solute dynamical states which can be further explained in terms of 
  solute-membrane interactions.
% order to automatically detect and parameterize an unknown number of
%  latent vector autoregressive dynamical modes. 
  %MRS: in above sentence: try to rephrase around the physical information obtained (mechanism or properties) rather than the mathematical route to it.
  %BJC2: gave a try

  %MRS: below - it's not the solute behavior, though, it's states identified by them model. Rephrase?
  %BJC2: I think this is better?
  We group the states identified by the IHMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane. This enhancement to our understanding of the diverse range 
  of influential solute-membrane interactions allows us to hypothesize design changes 
  to LLC monomers aimed towards controlling the rates of solute passage, thus improving 
  the selective performance of LLC membranes.

@@

  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms in 
  %BJC2: porous nanoscale to nanoporous? 
  %porous nanoscale 
  nanoporous materials and project complex diffusive behavior on
  long time scales.

@@

  % MRS: going to JPC, so emphasize the physical systems.
  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams.
  \begin{itemize}
    \item Desalination and boric acid removal from seawater.
    \item Organic micropollutants
    \item Hydraulic fracturing flowback water
    \item Breathable chemical warfare agent barriers
    \item High purity chemicals
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

  The ability to efficiently design novel membrane materials for selective separations
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables.
  %MRS: should explicitly point out that just calculating diffusion coefficients don't work for subdiffusive behavior, so
  %MRS need a different way to do it.
  %BJC2: added last two bullet points
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	\item In many systems, it's possible to achieve this by calculating diffusion 
  	constants.\cite{}
  	\item Unfortunately, for systems which exhibit complex subdiffusive dynamics on the
  	time scales accessible to MD simulation, one cannot calculate reliable diffusion 
  	constants since solute mean square displacements (MSDs) are non-linear.
  	\item Therefore, we are in need of a more flexible approach for modeling solute 
  	dynamics which makes less assumptions about the solute's average behavior.
  	%BJC2: I'm moving these towards the end
%  	\item There are two main challenges associated with this goal.
%  	%since complex and
%  	%diverse diffusive motion on time scales accessible to molecular simulations prevents
%  	%us from easily relating chemically dependent solute motion to experimental time scales.
%  	\item First, we need a simple way to identify and parameterize the fluctuations
%  	characteristic to dominant solute-membrane interactions.
%  	\item Second, we need a way to propagate this behavior far beyond the time scales
%  	afforded by MD.
  \end{itemize}

  %MRS: point out that LLC's are examples where there are lots of hopping/jumping (cite previous papers) and
  %MRS: so we need different approaches.
  %BJC2: added last bullet point to lead in to following paragraph
  In this work, we are particularly interested in the design of lyotropic liquid crystal (LLC)
  monomers, a class of amphiphilic molecules whose ordered self-assembled phases can be 
  cross-linked into mechanically strong membranes capable of highly selective separations.
  \begin{itemize}
  	\item Inverted hexagonal, H\textsubscript{II}, phase LLC membranes are characterized
  	by hexagonally packed, uniform-sized and straight pores, an ideal geometry for high 
  	throughput transport.
	\item Their pores are lined with the LLC monomer functional groups which can
	potentially be designed to interact with solutes in a chemically-specific manner.
	\item It would be extremely useful to experimentalists if we could connect LLC monomer
	design with measurable quantities.
	\item This goal is complicated by the non-Brownian hopping and trapping behavior
	induced by the membrane's diverse topology and inhomogeneous structure.~\cite{coscia_understanding_2019,coscia_chemically_2019}
  \end{itemize}

@@

  We aim to address two primary scientific 
  questions with this work. 
  First, we want
  to know if the IHMM algorithm can help researchers efficiently uncover underlying
  transport mechanisms which give rise to different solute dynamical behavior.
  Second, we want to predict macroscopic transport by using our model to generate 
  stochastic solute trajectories at much longer timescales than possible with MD simulations, 
  extending into timescale where diffusive behavior again occurs.

@@

@@

  %BJC2: I left section numbers off
  \subsubsection*{Extending Hidden Markov Modeling to an Unknown Number of States}

  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to the dynamical behavior which give rise to those types
  of observations. The probability of transitioning to a state based on the current
  state is mathematically defined in terms of an $n\times n$ transition probability
  matrix, $T$, where $n$ is the number of states. Unfortunately, standard HMMs 
  require $n$ to be known \textit{a priori}. One can partially overcome this by 
  testing a range of numbers of hidden states and determining which is the best 
  representation of the data.~\cite{pohle_selecting_2017}

@@

  \subsubsection*{Parameterizing the Hidden States}\label{method:var_params}

  We describe the dynamics of each state visited by solutes in our MD simulations using
  a first order vector autoregressive (VAR(1)) model. In general, a VAR($r$) process is characterized by a
  vector of observations in a time series that are linearly dependent on $r$ previous
  values of the time series vector:
  \begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  \label{eqn:var}
  \end{equation}
  Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  process is further characterized by a shift in the mean of each dimension by the
  vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  covariance, $\Sigma$. We limited our analysis to an autoregressive order of $r=1$.
  This means that we only parameterize $A_1$. To simplify notation, we will just
  call it $A$. We used a matrix-normal inverse-Wishart prior on parameters $A$ and 
  $\Sigma$ and a Gaussian prior on $\mathbf{c}$ in order to infer their 
  values.~\cite{fox_nonparametric_2009}

  \subsubsection*{Applying the IHMM to MD Trajectories}

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  % each entry in the $\mathbf{c}$ vector as well as the $A$ and $\Sigma$ matrices of Equation~\ref{eqn:var}. 
  We detected equilibration 
  of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

  %BJC2: I say 'applied' a lot. Diversity that eventually
  We carefully applied the IHMM algorithm in a way which takes advantage of the
  system's cylindrical symmetry. While there
  are a number of viable ways in which one could choose to analyze these time series,
  we chose to use a multi-step procedure that we believe adequately distinguishes 
  every type of distinct dynamical behavior exhibited by the solutes. The final
  parameterization of the model is in cylindrical coordinates since each pore is a
  cylinder where we expect solutes to exhibit radially symmetric dynamics. We 
  present a step-by-step demonstration of the procedure that follows in 
  Section~\ref{S-section:ihmm_procedure} of the Supporting Information.

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$. By working in 
  Cartesian coordinates, we avoid mathematical complexity introduced by cylindrical 
  coordinates while estimating the state sequence.

  We applied the IHMM to each of the 24 solute trajectories independently.
  Although the IHMM is capable of identifying an infinite number of states, 
  a Dirichlet process tends to exhibit a ``rich get richer" effect, favoring
  a fewer number of states.~\cite{dreyer_discovering_2011} By applying the algorithm to each trajectory 
  independently, we reduce the possibility of lumping together multiple 
  similar states which we would prefer to stay separated before clustering.
  Note that the states identified by the IHMM are heavily influenced by 
  the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. 
  In Section~\ref{S-section:prior_guesses} of the Supporting Information we 
  outline our method for choosing reliable prior parameters.

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences. We evaluated the convergence of the state sequence based on
  the convergence of the VAR parameters. The parameters of each state
  generally converge within 500--1000 iterations. However, the boundaries of 
  each state segment fluctuate which can lead to high variance in the $A$
  and $\Sigma$ parameters in segments with a relatively low number of emissions.
  In these cases, plateauing of the mean in each dimension tends to be a more reliable indicator of
  convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting Information).

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) due to the system's radial 
  symmetry. We converted the $x$ and $y$ center-of-mass coordinates to $r$. 
  We fixed the state sequence from the 3D parameterization and applied the 
  inference component of the IHMM procedure to the cylindrical trajectories 
  in order to estimate the VAR(1) parameters in terms of $r$ and $z$. Since
  we fixed the state sequence, and we are using conjugate priors, the VAR 
  parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
  Therefore, we only ran the inference procedure for 100 iterations. 
  % MRS: hmm. Here the new covariance parameters is redone in r - does there need to be any transformation applied? 
  % MRS: though might be OK, since x^2 + y^2 = r^2 so maybe covariance are additive if x and y uncorrelated (which they should be)

@@

  \subsection{Generating Stochastic IHMM Trajectory Realizations}\label{method:realizations}

@@

  \subsection{State Clustering}\label{method:clustering}

@@

  \subsubsection*{Choosing the Number of Clusters}

  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines. We used the silhouette 
  test in order to score the quality of clustering as a function of the number 
  of clusters chosen.~\cite{kaufman_finding_2009} For our data, the silhouette 
  test generally favors the lowest number of clusters possible (see 
  Section~\ref{S-section:nclusters} of the Supporting Information). However, 
  choosing too few clusters tends to not distinguish between visually obvious 
  differences in dynamic behavior. This results in finalized parameter sets 
  that are averages of distinct behavior which presents further problems with 
  the predictive modeling that we discuss later on. We aimed to maintain the 
  highest silhouette score, and thus lowest number of clusters, possible while
  verifying that visually distinct states stayed separated.

%  Based on these guidelines, we decided to 
%  group the $A$ and $\Sigma$ parameters into five clusters each and the radial
%  means into three clusters. Of the 75 distinct states allowed by this formulation,
%  the solutes in this study showed behavior from 32--36 distinct clusters. While
%  this may seem like a high number of states, many are sampled infrequently.
  %MRS: what is the consequence of being sampled infrequently.

  \subsubsection*{Obtaining Parameters of Clustered States}

  We remapped the state sequence based on the cluster assignments and parameterized
  the clusters using the inference component of the IHMM algorithm. 
  First, we modified the ($r$, $z$) solute trajectories so that they had a mean of zero,
  leaving only the fluctuations. We did this by subtracting the mean from each 
  same-state segment of the unclustered trajectory. We used the IHMM algorithm on this
  modified trajectory to infer the clustered state parameters by fixing the clustered 
  state sequence.

@@

  \subsection{Tools for using the Parameterized Model to Explore Mechanisms}\label{method:interactions}

@@

@@

  We applied the IHMM independently to all 24 trajectories of each of the four solutes studied.
  \begin{itemize}
    \item The number of states found for each trajectory varied between 5 and 30  %BJC2: need to see what these numbers actually are.
    \item In Figure~\ref{fig:rz_unclustered}, we show the state sequence determined by the
    IHMM for an example methanol trajectory.
    \item The model distinguished many of the states due to differences in their mean levels 
    while their fluctuations about those means are often similar. 
    \item In Section~\ref{section:mechanisms}, we use clustering to reduce the total state
    space for each solute.    
  \end{itemize}

  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{rz_unclustered_MET.pdf}
  \caption{The IHMM found 24 distinct VAR(1) states in this representative methanol trajectory.}\label{fig:rz_unclustered}
  \end{figure}

  \subsection{Reproducing MD Trajectories and MSDs with the IHMM}

  % MRS: maybe makes sense to talk a little about what features/parameters of the 
  % models the MSD is sensitive to?

  Before analyzing solute behavior characteristic to the states identified by the
  IHMM, it is important to verify whether the state dynamics are consistent with MD
  both qualitatively and quantitatively.
  \begin{itemize}
    \item Therefore, we generated stochastic realizations based on the parameters 
    of each trajectory as described in Section~\ref{method:realizations}.
  \end{itemize}

  As shown in Figure~\ref{fig:qualitative_unclustered}, realizations of our model 
  can produce trajectories that show qualitatively similar hopping and trapping 
  behavior to the MD trajectories.
  \begin{itemize}
    \item It is important that our model reproduces this behavior so that we can
    have confidence in its usefulness for quantitative predictions.
  \end{itemize}

  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{qualitative_unclustered_MET2.pdf}
  \caption{We can generate trajectories which bear qualitative resemblance to
  the MD trajectories. Solute trajectories generated by the IHMM (blue) show
  the same hopping and trapping behavior exhibited by MD (black).
  }\label{fig:qualitative_unclustered}
  \end{figure}

  \noindent From a quantitative standpoint, the model has two shortcomings.
  \begin{itemize}
  	\item First, in many cases, the MSD predicted by realizations of the IHMM 
  	under-predicts that of MD (see Figure~\ref{fig:unclustered_msds}).
  	\item Second, the shapes of the MSD curves at short time lags are very different.
  \end{itemize}
